\textcolor[rgb]{0,0,1}{Data-Driven Model Predictive Control using Random Forests for Energy and Thermal Comfort Optimization in Buildings}

### 2018-01-21 paper/final/Response_to_reviewers/Response_to_reviewers.tex

\documentclass{article}
\textwidth 6in%
\textheight 7.9in%
\oddsidemargin 1cm

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}%
\usepackage[dvips]{color}
\setcounter{MaxMatrixCols}{30}
\usepackage{natbib}

\begin{document}

\title{Authors' Response \\
\begin{Large}
	Data-Driven Model Predictive Control using Random Forests for Energy and Thermal Comfort Optimization in Buildings
\end{Large}
}

@@

The following are itemised lists of the authors' response to the Reviewers and to the Associate Editor.

\section{Response to Associate Editor}

\begin{enumerate}

	\item \textbf{Associate Editor:} \textit{The authors should address the points and clarifications in the reviews, and also specifically address the novelty of the paper with respect to prior work.}

@@

		\item add another section to show how the previous point simplifies using data-driven methods

	\end{itemize}

\item \textbf{Reviewer 1:} \textit{Another point is that in traditional state space model, one cannot measure everything except input and output in most cases. How could we obtain such training data if these data are not measurable.}

\textbf{Authors:} We thank the reviewer for this comment, indeed our previous presentation of the paper was unclear on this fundamental aspect. In our revision we widely illustrate in Section 2, how in traditional state-space models part of the state is measurable (i.e. room temperatures, etc) and part is not measurable (i.e. layers/windows temperatures, etc). In our approach, of course, $(\mathcal{X}^c, \mathcal{X}^d, \mathcal{Y})$ are ALL measurable variables, i.e. variables that are present in the historical data of a building. The unmeasurable variables of the traditional state-space models are not assumed to be measurable in this paper and are neglegted in the proposed models, yet our prediction works well since we compensate them via the regression tree/random forest approaches.

@@

\item \textbf{Reviewer 2:} \textit{The paper applies a recently developed concept, data predictive control (DPC) to building energy control. Models based on decision trees and random forests are applied. The efficacy of the approach appears well validated.}

\textbf{Authors:} We thank the Reviewer for the comment on our validation of the efficacy of our approach.

\item \textbf{Reviewer 2:} \textit{The title is a bit confusing as it's very general. It should be changed to be more specifically focusing on the applications in this paper.}

\textbf{Authors:} We agree with the Reviewer. In the beginning, we had chosen a more general title because we believe that our approach has the potential to be applied not only to building management, but also to other application domains. However, as in this paper we develop our techniques focusing and experimenting in building management (which is indeed a core topic of APEN) we changed the title in "Data-Driven Model Predictive Control using Random Forests for Energy and Thermal Comfort Optimization in Buildings".

@@

\item \textbf{Reviewer 2:} \textit{Also, novelties beyond what's in [9] - [12] should be clarified.}

\textbf{Authors:} In the introduction of the revised version, we widely discuss that in this paper we provide new results with respect to [9-12] (that now are [25,26,31,32]) and adapted them to be applicable to the building automation use-case presented in Section 6.

@@

\item \textbf{Reviewer 2:} \textit{In (1) and 2.1, the description of separation of variables is vague and needs to be more precise. E.g., the line below (1) does not make sense. It is not until later seeing the examples the meaning of this part becomes clearer. Such confusion persists up to and beyond (3). Mathematically, (2) and (3) cannot be simultaneously true. While I can see the authors trying to be descriptive, the descriptions lack rigor.}

\textbf{Authors:}  We agree with the reviewer. Following this comment, and also other comments, we completely re-wrote the corresponding Section (that now is Section 3) to make it more clear and complete to describe the DPC technical approach.

@@

\item \textbf{Reviewer 2:} \textit{The main optimization problems (4) and (7) are not clear in themselves. As you look ahead to do DPC, does each future time slot have a different linear model, depending on future non-manipulative states? Then, do you assume you know precisely all the future non-manipulative states? If so, can you justify this? Apparently, with a little error in predicting of the future states, a completely different linear model may be chosen. Will this change the performance significantly? Even with forests this is a major concern.}

\textbf{Authors:} We thank the reviewer for rising this important issue. Yes, we assume that at time $k$ we know the future non-manipulative states (namely external temperature, radiation, etc.) for the Model Predictive Control time horizon. In the revised version of this paper, to answer the reviewer comment, we account for disturbance uncertainties and show that, even with a large error in predicting the future states, the performance does not change significantly. More precisely, we:
	\begin{itemize}
		\item Cite a study that shows, despite considering uncertainties on weather forecast on a long horizon of 72 hours, that MPC outperforms the rule-based controllers in providing energy savings, and is in general quite close to the perfect forecast case. This is due mainly to 2 main features:
		\begin{itemize}
			\item weather forecast are in general quite accurate on a shot-term horizon (especially on the length of horizon that we usually consider for DPC, for example 6 hours in Section 4, 7 hours in Section 5 and 40 minutes in Section 6);
			\item MPC strategy computes an optimal input sequence over an horizon of finite length, and then apply only the first input. At the next step, with the new EXACT measurements of the system state and of the disturbance conditions, the algorithm is applied again. This is how MPC uses feedback law to correct modeling errors;
	\end{itemize}
		\item Provide a new case study in Section 6, where we inject a noise into the disturbance and show the robustness of our approach.
	\end{itemize}

@@

\item \textbf{Reviewer 2:} \textit{In Fig. 6b, what's the difference between blue and yellow, as they seem to completely overlap?}

\textbf{Authors:} Achin?

\item \textbf{Reviewer 2:} \textit{Also, what is new in section 5 compared with section 4? Even though a real house is claimed to be used, it's still tested using simulations and energyplus, like in the previous sections.}

\textbf{Authors:} We thank the reviewer for this comment, indeed the contribution of this section was not well emphasized. We improved this aspect at the beginning of the section. In particular, Section 6 provides the following differences with respect to Section 5:
\begin{itemize}
	\item In Section 5 we use the EnergyPlus model both to create the Random Forests model and to test the DPC performance.
	However, although the EnergyPlus model is quite accurate, it still provides simulated data, that differently from experimental data do not suffer of non-predictable imperfections such as random occupancy schedules, open/close windows, random light on/off switch, etc.. In section 6, instead, experimental data are used to build the Random Forests models, while the EnergyPlus model (created ad-hoc for the house) is ONLY used to make a fair comparison. Otherwise, using Random Forests models to compare DPC (that would have used the same model) and the bang-bang control would have been unfair;
	\item The optimization problems are different: in Section 5, we cut the peak of the power in a limited amount of time and require the temperature to be close to a desired value, while in Section 6 we minimize the energy usage while keeping temperature in a comfort range. Furthermore, in Section 6 we consider an on/off input, i.e. integer input variable, instead of a continuous one, showing that DPC can adapt to different problems;
	\item We introduce new simulations accounting for weather forecast uncertainties, other than the ones introduced with the experimental data, that show the robustness of DPC against disturbance errors.
\end{itemize}

\item \textbf{Reviewer 2:} \textit{A real world experiment on this house would be much more interesting.}

\textbf{Authors:} We thank the reviewer for this comment, indeed the contribution of this section was not well emphasized. We improved this aspect at the beginning of the section. In particular, Section 6 provides the following differences with respect to Section 5:

### 2018-01-21 paper/final/dpc.tex
Sec 3 - DPC reviewed, and partially updated response to reviewers

\section{DATA PREDICTIVE CONTROL}
\label{S:dpc}

\textcolor[rgb]{0,0,1}{The central idea behind DPC is to obtain control-oriented models using machine learning methodologies, and formulate the control problem in a way that Receding Horizon Control (RHC) can still be applied and the optimization problem can be solved efficiently.
In this paper, Regression Trees and Random Forests are considered (see Appendix ??? for details).}

\textcolor[rgb]{0,0,1}{To this aim, let an historical dataset $(\X,\Y)$, created using building measurements, be given to be used in the Regression Trees and Random Forests learning process.
Let $|(\X,\Y)| = n$, be the number of samples in the dataset.
The set of predictor variables (or features) $\X = \{(x(k),u(k),d(k))\}$, $\forall k = 1,\ldots,n$, is the set of samples $(x(k),u(k),d(k))$, measured at each time instant $k$, where $x(k)\in\mathbb{R}^{n_x}$ is the vector of the \emph{measurable} state variables, $u(k)\in\mathbb{R}^{n_u}$ is the vector of the input variables and $d(k)\in\mathbb{R}^{n_d}$ is the vector of the \emph{measurable and predictable} disturbance variables.
We recall that, as explained in Section \ref{secModelingIssuesExample}, here we do not use non-measurable states, such as temperature of wall layers, ceiling, floor, etc., since the data-driven methodology is able to compensate them.
As an example, the variables in $\X$ can be represented by the weather forecast, as the disturbances, the set-points information and building schedules, as the inputs, and the room temperatures and power consumption, as the states. 
The set of response variables $\Y = \{y(k)\},\ \forall k = 1,\ldots,n$, is the set of samples $y(k)\in\mathbb{R}^{n_y}$.
The response variables are represented by the state variables at the next time step, i.e. $y(k) = x(k+1)$.}

\textcolor[rgb]{0,0,1}{Our goal is to learn data-driven models, using Regression Trees and Random Forests, that relate the value of the response variables with the value of the predictor variables, and that can be used to set up an MPC problem.
}

\textcolor[rgb]{0,0,1}{To do this, we want to derive a model with a closed-form expression of the following form:
\begin{equation}\label{E:GenericModel}
	y(k)=f(x(k),u(k),d(k)),
\end{equation}
where $f$ is an analytical expression that describes the system's dynamics.
The issue that arises, when using Regression Trees and Random Forests, is that $f$ does not have a closed-form expression, hence \eqref{E:GenericModel} is not suitable for control and optimization.
One possibility could be to look for sub-optimal solutions using evolutionary algorithms \cite{Kusiak2009}, i.e. using heuristics.
In this section we provide a new methodology to determine a closed-form expression for $f$, by manipulating the dataset as shown starting from Section \ref{SS:sepvar}.}

\subsection{Dataset splitting}
\label{SS:sepvar}

\textcolor[rgb]{0,0,1}{The dataset splitting consists of partitioning the features set $\X$ into the sets $\X^c = \{u(k)\}\subset\X,\ \forall k=1,\ldots,n$, containing the control (or manipulated) variables, and $\X^d = \{(x(k),d(k))\}\subset\X,\ \forall k=1,\ldots,n$, containing the disturbance and state (or non-manipulated) variables.
The union of the two sets forms the full feature set of training, i.e. $\X \equiv \X^c \cup \X^d$.
%Our goal is to replace a model-based controller with a data-driven controller, where the latter depends only on the historical sensor data. 
%These measurements could directly represent one or more states in the model-based control framework. We denote these as responses $\Y \in \R$ for training, i.e. a $\Y$ represents a particular response and we can have separate models for multiple responses. We define the number of training samples by $|(\X,\Y)| = n$.
Using this splitting methodology, the training process is divided into two steps:
	\begin{enumerate}
		\item the trees and the forests are trained only using $\X^d$, which also eases the computational complexity.
			  It is important to note that besides external disturbance and the state measurements, $\X^d$ can also contains autoregressive terms of such variables;
		\item in each leaf of the tree (or in each leaf of each tree of the forest), a linear regression model is trained as a function of the inputs variables in $\X^c$.
		As we shall see in Section~\ref{SS:dpcrt} and \ref{SS:dpcrf}, this second step reduces the run-time control problem into a convex program.
	\end{enumerate}
This process is illustrated in Figure~\ref{F:dpc-sepvars} (left). For the sake of simplicity only trees are considered, but the same holds for the forests. The meaning of the variable $j$ in the linear model in the figure will be clear in the next section.
In Figure~\ref{F:dpc-sepvars} (right), the step 1 of the training process uses also the input variables, i.e. the classical training process without the dataset splitting.
The reason why this latter approach is not suitable for control will be discussed at the end of Section \ref{SS:dpcrt}.
}

\subsection{DPC-RT: DPC with Regression Trees}
\label{SS:dpcrt}
When the data has lots of features, which interact in complicated, nonlinear ways, assembling a single global model such as linear or polynomial regression can be difficult, and can lead to poor response predictions.
\textcolor[rgb]{0,0,1}{As discussed before, an approach to non-linear regression is to partition the data space into smaller regions, where the interactions are more manageable. }
This partition is repeated recursively until we finally get to small chunks of the data space where we can fit simple (eg. linear parametric) models. 
%Therefore, in \eqref{E:sepvars}, the global model $f$ has two parts: the recursive partition $g$, and a linear (and convex) model $h$ for each cell of the partition.

\begin{figure}[t!]
	\centering
	\includegraphics[width=20pc]{figures/dpc-sepvars.eps}
	\caption{Dataset splitting. \textit{Step 1:} Tree $\mathcal{T}_1$ is trained only with variables in $\X^d$ as the features. Tree $\mathcal{T}_2$ uses both the disturbances and states in $\X^d$, and the control variables in $\X_c$ for splitting. It is thus not computationally suitable for control. \textit{Step 2:} In the leaf $\ell_i$ of the trees, a linear regression model parametrized by $\beta_i$ is defined as a function only of the control variables.}
	\captionsetup{justification=centering}
	\label{F:dpc-sepvars}
\end{figure}

\textcolor[rgb]{0,0,1}{Now, in order to have a model that can be used for prediction in an MPC problem, with an horizon of arbitrary length $N$, our goal is to predict, at time $k$, the response $y$ for the next $N$ time steps, i.e. $y(k),\ldots,y(k+N)$, given the state measurements and the disturbance forecast at time $k$, i.e. $x(k)$ and $\tilde d(k),\ldots,\tilde d(k+N)$.
For the sake of simplicity and without any loss of generality, in the following, we consider only a scalar response variable, i.e. $y(k)\in\mathbb{R}$ (or $n_y = 1$).
However, as we will see in Sections \ref{S:casestudy} and \ref{S:realCaseStudy}, multiple trees (or forests) can be built for $n_y > 1$.}

\textcolor[rgb]{0,0,1}{Applying the dataset splitting, we build $N$ regression trees such that the response $y(k+j)$ of the $j^{th}$ tree depends upon $\delta_d$ autoregressive terms of the disturbances, and on the state at instant $k$ and its $\delta_x$ autoregressive terms:}

\textcolor[rgb]{0,0,1}{\begin{equation}\label{E:model_tree}
\T_j = \mathit{f}_{\mathrm{tree}} \left( d(k+j-\delta_d),\ldots,d(k+j),x(k-\delta_x),\ldots,x(k)  \right),\ j=0,\ldots,N,
\end{equation}}

\noindent \textcolor[rgb]{0,0,1}{with $\left( d(k+j-\delta_d),\ldots,d(k+j),x(k-\delta_x),\ldots,x(k)  \right)\in\X^d,\ k=1,\ldots,n$, and $\mathit{f}_{\mathrm{tree}}$ representing the tree structure.
Then, the linear models, as functions of the inputs variables $u\in\X^c$, in each leaf of the trees $\T_j,\ j=0,\ldots,N$, are defined as}

\textcolor[rgb]{0,0,1}{\begin{equation}\label{E:model_leaf}
y(k+j) =  \beta_j [1,u(k),\ldots,u(k+j) ]^T,\ j=0,\ldots,N.
\end{equation}}

\noindent \textcolor[rgb]{0,0,1}{Note that the coefficients $\beta_j$ are different for each leaf.
Equation~\eqref{E:model_leaf} implies that the prediction of response $y(k+j)$ at time $k$ is an affine combination of control inputs from time $k$ to $k+j$.
Thus, we have managed to linearize the original model dynamics via black-box modeling.
One of the advantages of this approach is that this two-step training is done off-line, so the time required to create the models does not affect the control execution in run-time.}

\textcolor[rgb]{0,0,1}{The problem now is: \emph{how to use this modeling framework to set up an MPC problem?} In run-time, given the disturbance forecast and the state measurements at time $k$, i.e. $\tilde d(k+j-\delta_d),\ldots,\tilde d(k+j),x(k-\delta_x),\ldots,x(k)$ given $k$, we can narrow down to a leaf of each tree in \eqref{E:model_tree} to retrieve the linear models in \eqref{E:model_leaf} for each step $j=0,\ldots,N$ of the horizon (see first part of the pseudo code given in Algorithm \ref{A:dpcrt}), as depicted in Figure \ref{F:dpc-sepvars} (left). These models are used to setup the MPC Problem \ref{P:dpcrt}.}

\textcolor[rgb]{0,0,1}{In particular, }when a new control action has to be determined, each tree (prediction step) contributes to a linear constraint in the optimization as a replacement for the state dynamics in the case of MPC. Thus, the MPC optimization problem, considering a quadratic cost ($Q \succeq 0, R \succeq 0$) function and the general case of multiple responses, i.e. $y(k)\in\mathbb{R}^{n_y},\ n_y\geq 1$, can be formulated as:

\begin{problem}\label{P:dpcrt}
	\begin{equation}
	\begin{aligned}
	& \underset{u_{k+j}}{\mathrm{minimize}} & & \sum_{j=0}^{N} y^\top_{k+j} Q y_{k+j} + u^\top_{k+j} R u_{k+j} + \lambda\epsilon_j \\
	& \mathrm{subject\ to }                 & & y_{k+j}     =   \beta_j [1,u^\top_{k},\ldots,u^\top_{k+j} ]^\top                   \\
	&                                       & & u_{k+j}    \in  \mathcal{U}                                                        \\
	&                                       & & |y_{k+j}|  \leq \bar{y}_{k+j} + \epsilon_j 										   \\
	&                                       & & \epsilon_j \geq  0							                                       \\
	&                                       & & j           =    0,\ldots,N.            									       \\
	\end{aligned}
	\label{E:dpcrt}
	\end{equation}
\end{problem}

\textcolor[rgb]{0,0,1}{Here, $Q \in \mathbb{R}^{n_y\times n_y}$ and $R \in \mathbb{R}^{n_u\times n_u}$ are weight matrices used to give more importance in the minimization to either $y$ or $u$.
This problem is solved as in the classical MPC formulation, i.e. at each time step $k=1,2,\ldots$ the optimal control sequence $u^*_k,\ldots,u^*_{k+N}$ is computed, and only the first input of the sequence is applied to the system: $u(k) = u^*_k$.
The slack variables $\epsilon_j$ are added to ensure recursive feasibility (since the equality constraint on $y$ is relaxed), i.e. to guarantee that Problem \ref{P:dpcrt} can provide a solution at each step $k=1,2,\ldots$.
$\lambda$ is a weight used to give more importance to bounds violation, than to the quadratic term, and viceversa. 
Of course, a different cost function can be chosen depending upon the application, i.e. it can be linear, nonlinear, etc., obviously changing the complexity of the problem.
In the current formulation, the data-driven control problem is reduced to a convex program which is very easy and efficient to solve.}
The pseudo code for DPC-RT is given in Algorithm~\ref{A:dpcrt}.
\textcolor[rgb]{0,0,1}{DPC algorithm is also graphically shown in Figure \ref{F:dpc-algo-rf} for the random forest case, but the same holds for single regression trees.}

\textcolor[rgb]{0,0,1}{\begin{remark}
	 If we did not consider the splitting procedure	of the dataset, and we also used input variables to learn the trees, as in the right side of Figure 1, the resulting
	 model would not have been suitable for control. This is because, since u is the variable we want to optimize, we do not know its value a priori to go through the trees and determine the correct leaves to find the models to use in the RHC problem.
\end{remark}}

\begin{algorithm}[t!]
	\caption{Data Predictive Control with Regression Trees}
	\label{A:dpcrt}
	\begin{algorithmic}[1]
		\State \textsc{Design Time (Off-Line)}
		\Procedure{Model Training using Dataset Splitting}{}
		\State Set $\X^c$ $\gets$ manipulated features
		\State Set $\X^d$ $\gets$ non-manipulated features
		\State Build $N$ predictive trees with $(\X^d,\Y)$ as defined in \eqref{E:model_tree}
		\ForAll{trees $\mathcal{T}_j$}
		\ForAll{the leaves $\ell_i$ of $\mathcal{T}_j$}
		\State Fit $ y(k+j) =  \beta_j \left[1,u(k),\ldots,u(k+j) \right]^T$ as in \eqref{E:model_leaf}
		\EndFor
		\EndFor
		\EndProcedure
		\State \textsc{Run Time}
		\Procedure{Predictive Control}{}
		\While{$k< k_{\mathrm{stop}}$}
		\ForAll{trees $\mathcal{T}_j$}
		\State Determine the leaf $\ell_i$ using $\X^d$ as in \eqref{E:model_tree}
		\State Obtain the linear model at $\ell_i$ trained in \eqref{E:model_leaf}
		\EndFor
		\State Solve optimization problem \ref{P:dpcrt} to determine optimal
		\State control actions $u^*_k,\ldots,u^*_{k+j}$
		\State Apply the first input $u(k)=u^*_k$
		\EndWhile
		\EndProcedure
	\end{algorithmic}
\end{algorithm}

\subsection{DPC-En: DPC with Ensemble Methods}
\label{SS:dpcrf}
Regression trees obtain good predictive accuracy in many domains. However, the models used in their leaves have some limitations regarding the kind of functions they are able to approximate.
The problem with trees is their high variance and that they can overfit the data easily.
A small change $\delta$ in the data can result in a different series of splits and thus violate the acceptable accuracy.
This is the price to be paid for estimating a tree-based structure from the data.

>>>>

\section{DATA PREDICTIVE CONTROL}

\label{S:dpc}

\textcolor[rgb]{0,0,1}{The central idea behind DPC is to obtain control-oriented models using machine learning methodologies on historical datasets of buildings and formulate the optimal control problem in a way that Receding Horizon Control (RHC) can be solved efficiently. Let an historical dataset $(\X,\Y)$ be given.
We define $\X = \{(x(k),u(k),d(k))\}_{k=1}^n$ the set of predictor variables (or features), i.e. the set of samples $(x(k),u(k),d(k))$ measured a time instants $k=1,\ldots,n$, where $x(k)\in\mathbb{R}^{n_x}$ is the vector of the \emph{measurable} state variables (e.g. the rooms temperatures and power consumption of the building), $u(k)\in\mathbb{R}^{n_u}$ is the vector of the \emph{measurable} input variables (e.g. the set-points and schedule information of the building) and $d(k)\in\mathbb{R}^{n_d}$ is the vector of the \emph{measurable and predictable} disturbance variables (e.g. weather historical data).
We define $\Y = \{y(k)\}_{k=1}^{n}$ the set of response variables, i.e. the set of measured samples $y(k)\in\mathbb{R}^{n_y}$ representing the variables we wish to predict with our model. In the most general case, when we want to predict all measured variables, the response variables are represented by the state variables at the next time step, i.e. $y(k) = x(k+1)$. However, in general, we only wish to predict a subset $y(k) = \bar x(k+1) \subset \mathbb{R}^{n_x}$ of variables, e.g. only room temperatures and power consumption.
Clearly, $|\X| = |\Y| = n$ is the number of time samples of the dataset.\\
We remark that, as explained in Section \ref{secModelingIssuesExample}, the dataset $(\X,\Y)$ does not contain non-measurable states such as temperature of wall layers, ceiling, floor, etc.: nevertheless, our simulative and experimental results show that our data-driven methodology is able to compensate the absence of such variables while providing excellent prediction accuracy.}

\textcolor[rgb]{0,0,1}{Our goal is to learn data-driven models, using Regression Trees and Random Forests, that relate the value of the response variables (possibly for a certain future time horizon) with the value of the predictor variables and can be used to set up an MPC problem. To this aim, in the simplest example of one time-step prediction, we need to derive a model with a closed-form expression of the following form:
\begin{equation}\label{E:GenericModel}
	y(k)=x(k+1)=f(x(k),u(k),d(k)).
\end{equation}
However classical Regression Tree and Random Forest algorithms do not provide a closed-form expression for $f$, hence they cannot be used for MPC.}

\textcolor[rgb]{0,0,1}{In the next section we provide a new methodology (DPC) that adapts the classical Regression Tree and Random Forest algorithms to determine a closed-form expression for $f$ that is efficiently applicable to MPC. For simplicity of presentation we first describe DPC using Regression Trees and then DPC using Random Forests.}

%==============================================================================================================

\subsection{DPC-RT: DPC with Regression Trees}

\label{SS:dpcrt}

\textcolor[rgb]{0,0,1}{In order to have a model that can be used for prediction in an MPC problem with a future horizon of arbitrary length $N$	 we need to predict, at time $k$, the response $y$ for the next $N$ time steps, i.e. $y(k),\ldots,y(k+N)$. For the sake of simplicity and without loss of generality we consider in this section only a scalar response variable, i.e. $y(k)\in\mathbb{R}$ ($n_y = 1$). We will show in Sections \ref{S:casestudy} and \ref{S:realCaseStudy} that multiple trees (or forests) can be easily built to account for the case when $n_y > 1$.}

\textcolor[rgb]{0,0,1}{Assume that, at time $k$, we want to predict the response variable at time $k+j$, i.e. $y(k+j),\ j\in \{1,\ldots,N\}$: when the data has lots of features interacting in complicated, nonlinear ways, assembling a single global model such as linear or polynomial regression can be difficult, and can lead to poor response predictions.}

\textcolor[rgb]{0,0,1}{As discussed before an approach to non-linear regression is to partition the dataset into smaller regions where the interactions are more manageable. This partition can be obtained by recursively splitting the dataset via an adaptation we propose to the Regression Tree algorithm (see Appendix 1 for details), and is repeated recursively until we finally get to small chunks of the dataset (i.e. the leaves of the regression tree) where we can fit simple (e.g. linear or affine) parametric models.}

%Therefore, in \eqref{E:sepvars}, the global model $f$ has two parts: the recursive partition $g$, and a linear (and convex) model $h$ for each cell of the partition.

\begin{figure}[t!]

	\centering

	\includegraphics[width=20pc]{figures/dpc-sepvars.eps}

	\caption{\textit{Step 1:} Tree $\mathcal{T}_j$ trained only with variables in $\X^d$ using adapted RT algorithm. Tree $\mathcal{T}'_j$ trained with variables in $\X^d$ and control variables in $\X^c$ using classical RT algorithm. \textit{Step 2:} In the leaf $\ell_i$ of the tree an affine model $\beta_{ij}$ is defined as a function only of the control variables dataset corresponding to leaf $\ell_i$.}

	\captionsetup{justification=centering}

	\label{F:dpc-sepvars}

\end{figure}

\textcolor[rgb]{0,0,1}{Our modification of the Regression Tree algorithm first partitions the features set $\X$ into the sets $\X^c = \{u(k)\}_{k=1}^n\subset\X$, containing the control (or manipulated) variables, and $\X^d = \{(x(k),d(k))\}_{k=1}^n\subset\X$, containing the disturbance and state (or non-manipulated) variables. The  union of the two non-intersecting sets forms the full feature set of training $\X \equiv \X^c \cup \X^d$.
%Our goal is to replace a model-based controller with a data-driven controller, where the latter depends only on the historical sensor data. 
%These measurements could directly represent one or more states in the model-based control framework. We denote these as responses $\Y \in \R$ for training, i.e. a $\Y$ represents a particular response and we can have separate models for multiple responses. We define the number of training samples by $|(\X,\Y)| = n$.
Then the training process is divided into the following two steps, which generate as output a tree $\mathcal{T}_j$ as illustrated in Figure~\ref{F:dpc-sepvars} (left):
\begin{enumerate}
	\item the splitting of the dataset only partitions $\X^d$ (see Appendix 1 and \cite{hastie2009elements} for technical details): this choice is necessary to make our model suitable for control, as we will clarify later on, and also reduces the computational complexity. To each leaf $\ell_i$ will correspond an equivalence class of data samples $\X^d_i$ of the partition of $\X^d$, with $\X^d_i\subset \X^d$: as a consequence at any time $k$, given $\delta_d$ autoregressive terms of the disturbances and $\delta_x$ autoregressive terms of the state, we can associate $( d(k+j-\delta_d),\ldots,d(k+j),x(k-\delta_x),\ldots,x(k) )$, to the corresponding leaf $\ell_i$
	\begin{equation}\label{E:model_tree}
		\ell_i = \mathit{f}_{\mathcal{T}_j} \left( d(k+j-\delta_d),\ldots,d(k+j),x(k-\delta_x),\ldots,x(k)  \right)
	\end{equation}
	if and only if $( d(k+j-\delta_d),\ldots,d(k+j),x(k-\delta_x),\ldots,x(k)) \in \X^d_i$.
	\item in each leaf $\ell_i$ of the tree $\mathcal{T}_j$ we derive, solving a convex program over the data samples in the partition $\X^d_i$ associated to leaf $\ell_i$, an affine function, with coefficients $\beta_{ij}$, that relates the response $y(k+j)$ to the previous control inputs in $\X^c$:
	\begin{equation}\label{E:model_leaf}
		y(k+j) =  \beta_{ij} [1,u(k),\ldots,u(k+j) ]^T
	\end{equation}
	Clearly the coefficients $\beta_{ij}$ are different for each leaf $\ell_i$ as they are derived from different sets of samples.
\end{enumerate}
Applying the above procedure for $j=0,\ldots,N$ we build $N$ regression trees $\mathcal{T}_0,\ldots,\mathcal{T}_N$: thus, we have managed to linearize the original model dynamics via black-box modeling. Our two-step training procedure, described by the pseudo code of lines 1-11 of Algorithm~\ref{A:dpcrt}, can be computed off-line: this is an important advantage because the time required to create the model does not affect the control execution in run-time.}

\textcolor[rgb]{0,0,1}{The next problem now is: \emph{how to use this modeling framework to set up an MPC problem?} We setup the MPC optimization problem in the general case of multiple responses, i.e. $y(k)\in\mathbb{R}^{n_y},\ n_y\geq 1$, as follows:}

\begin{problem}\label{P:dpcrt}
	\begin{equation}
		\begin{aligned}
		& \underset{u_{k+j},\epsilon_j}{\mathrm{minimize}} & & \sum_{j=0}^{N} y^\top_{k+j} Q y_{k+j} + u^\top_{k+j} R u_{k+j} + \lambda\epsilon_j \\
		& \mathrm{subject\ to }                 & & y_{k+j}     =   \beta_{ij} [1,u^\top_{k},\ldots,u^\top_{k+j} ]^\top                   \\
		&                                       & & u_{k+j}    \in  \mathcal{U}                                                        \\
		&                                       & & |y_{k+j}|  \leq \bar{y}_{k+j} + \epsilon_j 										   \\
		&                                       & & \epsilon_j \geq  0							                                       \\
		&                                       & & j           =    0,\ldots,N.            									       \\
		\end{aligned}
		\label{E:dpcrt}
	\end{equation}
\end{problem}

\textcolor[rgb]{0,0,1}{Here, $Q \succeq 0 \in \mathbb{R}^{n_y\times n_y}$ and $R \succeq 0 \in \mathbb{R}^{n_u\times n_u}$ are weight matrices used to tradeoff the importance we provide in the minimization of $y$ versus $u$. The slack variables $\epsilon_j$ are added to ensure recursive feasibility: we relax the equality constraint on $y$ allowing violation up to $\epsilon_j$ to guarantee that Problem \ref{P:dpcrt} can provide a solution at each step $k \geq 0$. The weight $\lambda$ is then used to tradeoff the importance we provide in the optimal solution to the constraints violation versus the quadratic term. For example in the following sections we will use $\epsilon_j$ and $\lambda$ to define our tolerance to the violation of prescribed bounds by the rooms' temperature.\\
Clearly different cost functions can be chosen depending upon the application, i.e. they can be linear, nonlinear, etc., obviously changing the complexity of the problem. In the current formulation, the data-driven control problem, is reduced to a convex program which is very easy and efficient to solve in C, C++, Python and Matlab, and thus is easily integrable in SCADA systems. Indeed Problem \ref{P:dpcrt} is solved as in the classical MPC formulation, i.e. at each time step $k=1,2,\ldots$ the optimal control sequence $u^*_k,\ldots,u^*_{k+N}$ is computed, and only the first input of the sequence is applied to the system: $u(k) = u^*_k$.\\
Note that each tree $\T_j$ contributes to Problem  \ref{P:dpcrt} with the linear constraint $y_{k+j} = \beta_{ij} [1,u^\top_{k},\ldots,u^\top_{k+j} ]^\top$ as a replacement for the state dynamics in the classical MPC formulation. As a consequence, when solving Problem \ref{P:dpcrt} at time $k$, we need to determine the affine functions $\beta_{ij}$: it follows by Equation \eqref{E:model_tree} that, to determine $\beta_{ij}$ at time $k$, the knowledge of the state and disturbance measurements $\left( d(k+j-\delta_d),\ldots,d(k+j),x(k-\delta_x),\ldots,x(k) \right)$ is needed, however, values of $d(k+j-\delta_d),\ldots,d(k+j),\ j=0,\ldots,N$ are not available at time $k$, so we need to use disturbance forecast $\tilde d(k+N),\ldots,\tilde d(k)$. Using this information we can narrow down to a leaf in each tree using \eqref{E:model_tree} and thus retrieve the affine model with $\beta_{ij}$ in \eqref{E:model_leaf} for each step $j=0,\ldots,N$ of the horizon. The run-time solution of Problem \ref{P:dpcrt} illustrated above is described by the pseudo code of lines 12-23 of Algorithm~\ref{A:dpcrt}.}

\textcolor[rgb]{0,0,1}{\begin{remark}
It is now easy to understand that using the classical Regression Tree algorithm, e.g. using also the input variable $u$ in the data splitting procedure to create the trees as in the right side of Figure~\ref{F:dpc-sepvars}, the resulting model would not be suitable for control. Indeed, since $u$ is the variable we want to optimize in Problem \ref{P:dpcrt}, at time $k$ we have still not chosen its value at times $k, \ldots, k+N$: as a consequence the affine functions \emph{parameters }$\beta_{ij}$ needed to set up Problem \ref{P:dpcrt} cannot be determined at time $k$.
\end{remark}}

\textcolor[rgb]{0,0,1}{The pseudo code for the whole DPC-RT procedure (i.e. Off-line and Run-time) is given in Algorithm~\ref{A:dpcrt}. Our procedure is also graphically described in Figure \ref{F:dpc-algo-rf} for the Random Forest case, providing a good intuition also for the Regression Tree case.}

\textcolor[rgb]{0,0,1}{\begin{algorithm}[ht!]
	\caption{Data Predictive Control with Regression Trees}
	\label{A:dpcrt}
	\begin{algorithmic}[1]
		\State \textsc{Design Time (Off-Line)}
		\Procedure{Model Training using Dataset Splitting}{}
		\State Set $\X^c$ $\gets$ manipulated features
		\State Set $\X^d$ $\gets$ non-manipulated features
		\State Build $N$ predictive trees with $(\X^d,\Y)$ as defined in \eqref{E:model_tree}
		\ForAll{trees $\mathcal{T}_j$}
		\ForAll{leaves $\ell_i$ of $\mathcal{T}_j$}
		\State Fit $ y(k+j) =  \beta_{ij} \left[1,u(k),\ldots,u(k+j) \right]^T$ as in \eqref{E:model_leaf}
		\EndFor
		\EndFor
		\EndProcedure
		\State \textsc{Run Time}
		\Procedure{Predictive Control}{}
		\While{$k< k_{\mathrm{stop}}$}
		\ForAll{trees $\mathcal{T}_j$}
		\State Determine the leaf $\ell_i$ using $\X^d$ as in \eqref{E:model_tree}
		\State Obtain the linear model at $\ell_i$ trained in \eqref{E:model_leaf}
		\EndFor
		\State Solve Problem \ref{P:dpcrt} to determine optimal
		\State control actions $u^*_k,\ldots,u^*_{k+j}$
		\State Apply the first input $u(k)=u^*_k$
		\EndWhile
		\EndProcedure
	\end{algorithmic}
\end{algorithm}}

%==============================================================================================================

\subsection{DPC-En: DPC with Ensemble Methods}

\label{SS:dpcrf}

\textcolor[rgb]{0,0,1}{Regression trees obtain good predictive accuracy in many domains. However, the models used in their leaves have some limitations regarding the kind of functions they are able to approximate. The problem with trees is their high variance and that they can overfit the data easily, and a small change in the data can result in a different series of splits and thus affect the prediction accuracy. This is the price to be paid for estimating a tree-based structure from the data.}
